Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-09-05 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
fix latex issues in Mad Thesis

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 37 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica for "buying me time" to work on my studies.}

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupanies.  A common height for a residential occupanices is 2.4~m in height.  With a wall height of 2.4~m, these experiments will use small fires with mean flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

\Subsection{The Fire Investigators' Challenge}

For the moment lets step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 in the morning on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occured in area of older townhomes.  The socio-economics of the neighboorhood are such the the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene, "all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The fornt door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and chaild were rescued from a bedroom upstairs".  You thank the fire officier and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his work place.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit, townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  The is some thermal damage and smoke deposits on the from of the siding on the porch.  There outer pane of the living room window is for the most part still intact. you notice that three of the four windows on the second floor have been removed.  On the "D side" of the townhome there are no openings and no smoke or fire damage is present.  On the rear or "Side C"
of the house the door and windows are intact but open.  There is no somke stains or thermal damage around any of the openings on the rear. No sign of smoke alarms in the home

You take some notes and photograph the exterior of the structure, the fire department is not repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage.  Looking to your right, two chairs infront of the living room window have some thermal damage.  There are a clear horizontal line of demarcation the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the "D wall" of the living room was a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is not thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an uphostered sofa and chair.  The wall behind the sofa and chair has extensive thermal danage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  Their is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  THe fire damage doe not extend beyond the top of the stairs.  there is smoke damage in the upstairs hallway an in the front bedrooms. THE FURTHER YOU MOVE FROM THE stairs the less damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focussing on the area with the most damage, the area with the sofa and chair.  As you are looking toward 'b wall' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  The corner formed by the wall behind and the wall of the stairway has a "V pattern" burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The wire from the heater extends toward the wall outlet behind the sofa.
You have no developed your first hypothesis about the origin of the fire.  A radiant style, electric heater was turned on and left until it ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine of the heater switch was in the "on" position.  No you need to start testing your hypothesis.  Is everything consistant with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the rightside back cushion of the sofa.
the damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee looks like a arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the soaf frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further is appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair clearly has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not be moved.  The cushion was heated from the top down.  Perhaps the constrion of the chair slowed the flame spread through the chair from the bottom up, if the heatere was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews whatelse can you do to eliminate the inconsistancies?   We will reexamine the in Chapter XX.

@@

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Associaand tion, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 100~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11~\% It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a "square-wave".

The gasoline pan fire has was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a "bell shape"  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignitied and burned to completion, the polyurthane foam also generated "bell shaped" heat release rate curve.  Since the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~s interval bounding the peak heat release rate from each of the ten experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 47~\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

@@

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silcate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of it's thermal properties or in terms of its physical change as a result to exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, commonly this process is known as calcination.  Schroder exposed samples of gyspum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction  add Schroder.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dyhydrated gypsum to map fire patterns.

@@

As defined in NFPA 921, a fire pattern is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA_921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA_921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatibility of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those experiments provide valuable insight, it would be of valauble to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests.

@@

For the ignition tests, an electric spark ignitor was positioned over the center of the horizontal samples to serve as a pilot.  Combustion products and dilution air were extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate was kept constant during testing. The sample was mounted on a load cell to determine mass loss rate during a test, and smoke obscuration was measured using a laser light source. The gas flow rate in the exhaust duct was calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate was calculated from the gas concentration and mass flow measurements.

@@

Each apparatus provided similar but different thermal exposure conditions and each test method has a specific parameter or parameters that they were designed to measure for. Currently there is no specific protocol for a fire model user to follow to develop input data.  Nor is there a "standard" catalog of fire model inputs for a given material.  As a result the literature is sourced for information and users may to conduct experiments, such as those above, in order to develop values or ranges of values which could be used in the calculations in Chapters 8 and 9 or to evaluate the validity of the calculations in those chapters.

@@

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA_921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell or approximately 0.003~m$^2$.

@@

There are also a number of qualitative terms from NFPA~921 that were used to describe features of the patterns.  For example, "clean burn" to describe a lack of soot within a fire pattern area or "inverted cone" to describe the shape of the pattern~\cite{NFPA_921}.

@@

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each patterns is similar to the width of the burner. Moving up from the burner, about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and combusts, the plume and the pattern expands.  As the flames move up the wall the width tends to increase until about half of the peak height of the pattern, then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame is reduced.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticable bias to one side or the other can be seen in four of the ten experiments.

@@

In this chapter the application of emprical algebraic fire characterizations will be examined. Two different types of applications will be used.  In the first case relationships between the energy generated by a given fire source and the potential to transfer enough heat to generate thermal damage on a target, in this case gypsum board, will be examined.  In the second case, the height of the thermal damage on the gypsum board will be examined.  
Applications of algorithms such as these are presented in current fire investigaton textbooks~\cite{Icove:2013}.

@@

The first algorithm assumes that the heat release rate is emitted from a point source in a uniform manner.  Therefore for any given radii a sphere of nearly uniform heat flux will be assumed. A schematic of this arrangement is shown in Figure~\ref{PointSourceHF} and the equation is presented below. Babrauskas applied this concept in his paper, "Will the Second Item Ignite"~\cite{Babrauskas:1981}.  The idea being that if you knew the threshold heat flux that a material needed to be exposed to for the item to ignite.  Then given a fire with a specified HRR and radiant fraction, one could determine the range of distances from the fire where the material could recieve enough energy to ignite.  Radiation is assumed to be the only means of heat transfer.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/PointSourceHF} \\
	\caption[Schematic of heat flux estimation using the point source assumption]
	\label{PointSourceHF}
\end{figure}

@@

Geometry – the domain of the model was cubic with 2.4~ m on a side.  The resulting number computational cells for each case ranged from 27,000 for the 80~mm case to 1,728,000 for the 20~mm cases.   An isometric view of the model outlined by the domain is shown in Figure     .  The face of the wall is located 1.35~m from the front of the domain.  
The wall was modelled with material properties for gypsum board.
Thermal conductivity = 0.28
Specific Heat = 1.00
Density = 582

Instruments – FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  This devices were positioned  in locations similar to the locations used in the physical experiments, in the plume region and on the wall.   A radiative heat flux gauge was also modelled to provide data at the 0.95 m from the centerline of the burner and 1.0 m above the top of the burner.  
In addition, FDS features devices that can not be implemented such as a device, which can be used to examine the heat release rate per unit length.  In this case, this device was used to determine flame height based on that measure.     
Fire 
The burner was modeled as an obstruction, with the same physical size as the burner.  It was modelled with a steady state heat release rate of 888.89 kW/sq m.   The chosen fuel reaction was `methane’ since the majority <<<check percentage>> of the natural gas used in the laboratory experiments was methane.

>>>>

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 37 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica for ``buying me time'' to work on my studies.}

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m in height.  With a wall height of 2.4~m, these experiments will use small fires with mean flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

\subsection{The Fire Investigators' Challenge}

For the moment lets step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 in the morning on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older townhomes.  The socio-economics of the neighborhood are such the the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene, ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.''  You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his work place.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit, townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  The is some thermal damage and smoke deposits on the from of the siding on the porch.  There outer pane of the living room window is for the most part still intact. you notice that three of the four windows on the second floor have been removed.  On the ``D side'' of the townhome there are no openings and no smoke or fire damage is present.  On the rear or ``Side C'' of the house the door and windows are intact but open.  There is no smoke stains or thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is not repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage.  Looking to your right, two chairs infront of the living room window have some thermal damage.  There are a clear horizontal line of demarcation the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room was a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is not thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  Their is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The fire damage doe not extend beyond the top of the stairs.  there is smoke damage in the upstairs hallway an in the front bedrooms. The further you move from the stairs the less damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focusing on the area with the most damage, the area with the sofa and chair.  As you are looking toward the ``B wall'' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  The corner formed by the wall behind and the wall of the stairway has a ``V pattern'' burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The wire from the heater extends toward the wall outlet behind the sofa.

You have no developed your first hypothesis about the origin of the fire.  A radiant style, electric heater was turned on and left until it ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine of the heater switch was in the ``on'' position.  No you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the right side back cushion of the sofa. The damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee looks like a arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the sofa frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further is appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair clearly has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not be moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews what else can you do to eliminate the inconsistencies?   We will reexamine the in Chapter XX.

@@

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 100~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11~\% It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a ``square-wave''.

The gasoline pan fire has was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a ``bell shape''  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated ``bell shaped'' heat release rate curve.  Since the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~s interval bounding the peak heat release rate from each of the ten experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 47~\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

@@

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of it's thermal properties or in terms of its physical change as a result to exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, commonly this process is known as calcination.  Schroder exposed samples of gypsum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction  add Schroder.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dyhydrated gypsum to map fire patterns.

@@

As defined in NFPA 921, a fire pattern is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA:921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA:921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those experiments provide valuable insight, it would be of valuable to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests.

@@

For the ignition tests, an electric spark igniter was positioned over the center of the horizontal samples to serve as a pilot.  Combustion products and dilution air were extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate was kept constant during testing. The sample was mounted on a load cell to determine mass loss rate during a test, and smoke obscuration was measured using a laser light source. The gas flow rate in the exhaust duct was calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate was calculated from the gas concentration and mass flow measurements.

@@

Each apparatus provided similar but different thermal exposure conditions and each test method has a specific parameter or parameters that they were designed to measure for. Currently there is no specific protocol for a fire model user to follow to develop input data.  Nor is there a ``standard'' catalog of fire model inputs for a given material.  As a result the literature is sourced for information and users may to conduct experiments, such as those above, in order to develop values or ranges of values which could be used in the calculations in Chapters 8 and 9 or to evaluate the validity of the calculations in those chapters.

@@

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA:921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell or approximately 0.003~m$^2$.

@@

There are also a number of qualitative terms from NFPA~921 that were used to describe features of the patterns.  For example, ``clean burn'' to describe a lack of soot within a fire pattern area or ``inverted cone'' to describe the shape of the pattern~\cite{NFPA:921}.

@@

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each patterns is similar to the width of the burner. Moving up from the burner, about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and combusts, the plume and the pattern expands.  As the flames move up the wall the width tends to increase until about half of the peak height of the pattern, then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame is reduced.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticeable bias to one side or the other can be seen in four of the ten experiments.

@@

In this chapter the application of empirical algebraic fire characterizations will be examined. Two different types of applications will be used.  In the first case relationships between the energy generated by a given fire source and the potential to transfer enough heat to generate thermal damage on a target, in this case gypsum board, will be examined.  In the second case, the height of the thermal damage on the gypsum board will be examined. Applications of algorithms such as these are presented in current fire investigation textbooks~\cite{Icove:2013}.

@@

The first algorithm assumes that the heat release rate is emitted from a point source in a uniform manner.  Therefore for any given radii a sphere of nearly uniform heat flux will be assumed. A schematic of this arrangement is shown in Figure~\ref{PointSourceHF} and the equation is presented below. Babrauskas applied this concept in his paper, ``Will the Second Item Ignite''~\cite{Babrauskas:1981}.  The idea being that if you knew the threshold heat flux that a material needed to be exposed to for the item to ignite.  Then given a fire with a specified HRR and radiant fraction, one could determine the range of distances from the fire where the material could receive enough energy to ignite.  Radiation is assumed to be the only means of heat transfer.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/PointSourceHF}
	\caption{Schematic of heat flux estimation using the point source assumption}
	\label{PointSourceHF}
\end{figure}

@@

Geometry – the domain of the model was cubic with 2.4~ m on a side.  The resulting number computational cells for each case ranged from 27,000 for the 80~mm case to 1,728,000 for the 20~mm cases.   An isometric view of the model outlined by the domain is shown in Figure     .  The face of the wall is located 1.35~m from the front of the domain.  
The wall was modeled with material properties for gypsum board.
Thermal conductivity = 0.28
Specific Heat = 1.00
Density = 582

Instruments – FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  This devices were positioned  in locations similar to the locations used in the physical experiments, in the plume region and on the wall.   A radiative heat flux gauge was also modelled to provide data at the 0.95 m from the centerline of the burner and 1.0 m above the top of the burner.  
In addition, FDS features devices that can not be implemented such as a device, which can be used to examine the heat release rate per unit length.  In this case, this device was used to determine flame height based on that measure.     
Fire 
The burner was modeled as an obstruction, with the same physical size as the burner.  It was modeled with a steady state heat release rate of 888.89 kW/sq m.   The chosen fuel reaction was `methane’ since the majority <<<check percentage>> of the natural gas used in the laboratory experiments was methane.
